
---
title: "Wide vs Long Data in R"
author: "Jerson Garcia"
output:
  html_document:
    theme: readable
    highlight: zenburn
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T, message=F, warning=F)
```

# Introduction

Data can be organized in different ways.

Two common formats are **wide data** and **long data**.

Wide data usually has one row for each person or observation and several columns for different measurements.

Long data usually has more rows, but fewer columns. The measurements are stacked into one column.

This tutorial shows how to move between wide and long data using `pivot_longer()` and `pivot_wider()`.

# Packages

We will use `dplyr` and `tidyr`.

```{r}
library(dplyr)
library(tidyr)
```

# Wide Data

Here is a small dataset with student grades.

```{r}
grades <- data.frame(
  student=c("Alex","Sam","Taylor"),
  math=c(90,85,88),
  science=c(92,81,95),
  english=c(87,89,91)
)

grades
```

This is **wide data**.

Each student has one row, and each subject has its own column.

```text
student   math   science   english
Alex       90       92        87
Sam        85       81        89
Taylor     88       95        91
```

# Converting Wide Data to Long Data

We can use `pivot_longer()` to stack the subject columns.

```{r}
grades.long <- 
  grades %>% 
  pivot_longer(
    cols=c(math,science,english),
    names_to="subject",
    values_to="score"
  )

grades.long
```

Now each student has several rows.

The subject names are stored in one column called `subject`, and the grades are stored in one column called `score`.

# What Changed?

The original data had:

```{r}
dim(grades)
```

The long data has:

```{r}
dim(grades.long)
```

The number of rows increased because the three subject columns were stacked.

# Looking at One Student

The long format makes it easy to filter one student.

```{r}
grades.long %>% 
  filter(student == "Alex")
```

Now all of Alex's scores are stored in separate rows.

# Why Long Data Can Be Useful

Long data is useful when we want to summarize values by category.

For example, we can calculate the average score for each subject.

```{r}
grades.long %>% 
  group_by(subject) %>% 
  summarize(avg.score=mean(score))
```

This is easier because all of the scores are stored in the same `score` column.

# Another Summary

We can also find the highest score for each subject.

```{r}
grades.long %>% 
  group_by(subject) %>% 
  summarize(high.score=max(score))
```

# Using pivot_longer()

The basic structure is:

```{r eval=FALSE}
data %>% 
  pivot_longer(
    cols=columns_to_stack,
    names_to="new_name_column",
    values_to="new_value_column"
  )
```

In our example:

```{r eval=FALSE}
grades %>% 
  pivot_longer(
    cols=c(math,science,english),
    names_to="subject",
    values_to="score"
  )
```

The column names:

```text
math
science
english
```

become values inside the `subject` column.

# Selecting Columns Another Way

We do not always have to type every column name.

If all columns except `student` should be reshaped, we can use:

```{r}
grades.long2 <- 
  grades %>% 
  pivot_longer(
    cols=-student,
    names_to="subject",
    values_to="score"
  )

grades.long2
```

`-student` means use every column except `student`.

# Converting Long Data Back to Wide

We can reverse the process with `pivot_wider()`.

```{r}
grades.wide <- 
  grades.long %>% 
  pivot_wider(
    names_from=subject,
    values_from=score
  )

grades.wide
```

Now each subject is back in its own column.

# Using pivot_wider()

The basic structure is:

```{r eval=FALSE}
data %>% 
  pivot_wider(
    names_from=category_column,
    values_from=value_column
  )
```

For our grades:

```{r eval=FALSE}
grades.long %>% 
  pivot_wider(
    names_from=subject,
    values_from=score
  )
```

# Wide vs Long

Wide data:

```{r}
grades
```

Long data:

```{r}
grades.long
```

Wide format is useful when we want to see several measurements side by side.

Long format is useful when we want to group, summarize, or graph one measurement across categories.

# A Common Mistake

If we accidentally include the student column inside pivot_longer(), R will try to combine the character student names with the numeric scores. Since these are different data types, this can cause an error.

For example:

```{r eval=FALSE}
grades %>% 
  pivot_longer(
    cols=everything(),
    names_to="variable",
    values_to="value"
  )
```

Identifier columns like student should normally stay separate.


A better choice is:

```{r eval=FALSE}
grades %>% 
  pivot_longer(
    cols=-student,
    names_to="subject",
    values_to="score"
  )
```

# Small Example with Years

The same idea can be used for other types of data.

```{r}
sales <- data.frame(
  store=c("A","B","C"),
  `2024`=c(100,120,90),
  `2025`=c(110,135,105),
  `2026`=c(125,140,115),
   check.names=FALSE
)

sales
```

This dataset is wide because each year has its own column.

Convert it to long format:

```{r}
sales.long <- 
  sales %>% 
  pivot_longer(
    cols=-store,
    names_to="year",
    values_to="sales"
  )

sales.long
```

Now it is easier to compare values by year.

```{r}
sales.long %>% 
  group_by(year) %>% 
  summarize(avg.sales=mean(sales))
```

# Quick Reference

Wide to long:

```{r eval=FALSE}
data %>% 
  pivot_longer(
    cols=-id,
    names_to="category",
    values_to="value"
  )
```

Long to wide:

```{r eval=FALSE}
data %>% 
  pivot_wider(
    names_from=category,
    values_from=value
  )
```

Useful functions:

```{r eval=FALSE}
pivot_longer()

pivot_wider()

group_by()

summarize()

filter()
```

# Conclusion

Wide and long data contain the same information, but the information is organized differently.

`pivot_longer()` stacks columns into rows.

`pivot_wider()` spreads values back into separate columns.

Knowing how to move between the two formats makes it easier to prepare data for summaries, graphs, and other analysis.



